Intrinsic Verification of Parsers and Formal Grammar Theory in Dependent Lambek Calculus (Extended Version)

### 2025-04-21 paper/paper.tex
more tweaks. One more section left

To show the feasibility of our design, we have implemented \theoryabbv
as a shallowly embedded domain-specific language in the Cubical Agda
proof assistant \cite{VezzosiMortbergAbel2019}. We have implemented many example
grammars and parsers in our system including regular expressions,
non-deterministic and deterministic automata, as well as some example
context-free grammars and parsers based on $\LL{1}$ and $\LR{1}$
automata. Throughout this paper \Agda~ will mark
results that are mechanized in our Agda development and provide a link to their
implementation.

@@

\begin{axiom}
  \label{ax:string-top}
   $\lambda s. \mathsf{read}(!(s)) \equiv \lambda
s. s$ where $!$ is the unique function $\ltonl(\StringGram\lto \top)$. That is, $\StringGram$ is strongly equivalent to $\top$.
\end{axiom}

@@

\section{Formal Grammar Theory in \theoryname}
\label{sec:applications}
This section explores the applications of \theoryabbv to formal
grammar theory. We demonstrate that several classical notions and constructions
integral to the theory of formal languages are faithfully represented
in \theoryname. By encoding well-established formal grammar concepts, we ensure
that our framework remains grounded in the
foundational principles of formal language theory while opening the door to
compositional formal verification of parsers.

@@

A formal grammar $\A$ is ambiguous if there are multiple parse trees
for the \emph{same} string $\w$.  For example, $\a \oplus \a$ is
ambiguous because there are two parses of $\stringquote{a}$,
constructed using $\inl$ and $\inr$. On the other hand, a formal
grammar is unambiguous when there is at most one parse tree for any
input string. We can capture this notion as a type in \theoryabbv in a
clever way:
\begin{definition}
  \label{def:unambig}
  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
  then $\f \equiv \g$.
\end{definition}
\cref{def:unambig} can be read more intuitively as stating that $\A$
is unambiguous if there is at most one way to transform parses of any
other grammar $\B$ into parses of $\A$. This notion of an unambiguous
type is the analog for linear types of the definition of a (homotopy)
\emph{proposition} in the terminology of homotopy type
theory \cite{hottbook}. The most basic unambiguous types are $\top$ and
$0$, and in a system of classical logic all unambiguous types would
have to be equivalent to one of these, but with our axioms we can show
also that $\I$ and literals $\literal c$ are unambiguous. To see this, first, we
establish two useful properties of unambiguity.
\begin{lemma}[\Agda]
  \label{lem:retract}
  If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
\end{lemma}
\begin{lemma}[\Agda]
  \label{lem:unambig-sum}
  As a consequence of \cref{cor:binary-mono}, if a binary disjunction $A \oplus B$ is unambiguous then
  $A$ and $B$ are each unambiguous.
\end{lemma}
From \cref{lem:retract}, we can prove that $\StringGram$ is unambiguous,
since it is a retract of $\top$. In fact, observe that if $\A$ is a
retract of $\B$ and $\B$ is unambiguous, then in fact $\A$ and $\B$
are strongly equivalent, as the equation $\lambda b. f(g(b)) \equiv
\lambda b. b$ follows because $\B$ is unambiguous. Therefore
$\StringGram$ is also strongly equivalent to $\top$. Next, since $\StringGram$ is
defined as a Kleene star, we can easily show that $\StringGram \cong \I
\oplus \CharGram \oplus (\CharGram \otimes \CharGram \otimes \StringGram)$. Then by
\cref{lem:unambig-sum}, we have that $\I$ and
$\CharGram$ are unambiguous as well. Using the finiteness of the alphabet
$\Sigma$ and the unambiguity of $\CharGram$,
we have that each literal $\literal c$ is likewise unambiguous.

@@

In this section, we describe how to construct an intrinsically
verified parser for regular expressions by compiling it to an NFA and
then a DFA. That is, for each regular expression $A$, we construct an
NFA $N(A)$ and a corresponding DFA $D(A)$ such that $A$ is strongly
equivalent to the traces of $N(A)$ and weakly equivalent to the
accepting traces of $D(A)$. Then we can easily construct a parser for
traces of $D(A)$ and apply \cref{lem:wk-eqv-parse} to get
a verified regular expression parser.

@@

Because DFAs are deterministic, we are able to prove that their type
of traces are unambiguous and define a parser for them. In particular
we show that for any start state $s$, ${\LinSigTy {b} {\Bool}
  \Trace_{D}~s~b}$ is a retract of $\StringGram$ and apply \cref{lem:retract} to
derive unambiguity. That is, we first
construct a function $\pparse_{D}$ which is a parser for
$\Trace_D~{s}~{true}$ with $\Trace_D~{s}~{false}$ being the disjoint
type used, and disjointness follows from the
unambiguity of $\LinSigTy {b} {\Bool} {\Trace_{D}~s~b}$ by \cref{lem:unambig-sum}.

@@

The Dyck language is an example of an $\textrm{LL}(0)$ language, one
that can be parsed top-down with no
lookahead \cite{rosenkrantz_properties_1970}. This means we can
implement it simply as an \emph{infinite} state deterministic
automaton, in \Cref{fig:DyckAutomaton}. Here the state is a ``stack''
counting how many open parentheses have been seen so far. Functions
$parse_{M}$ and $print_{M}$ for this automaton can be defined
analogously to those for DFAs, and so $\LinSigTy {s}
{M .states} {\LinSigTy {b}{Bool} {Trace_{M}~s~b}}$ is likewise
unambiguous.

@@

This reification operation on functions $\StringSem \to U$ is incredibly expressive,
as it sidesteps our linear typing connectives and utilizes the whole of nonlinear
dependent type theory to define a grammar. For example, given a
Turing machine $T$ one may
define a non-linear predicate $accepts : \StringSem \to U$ that encodes that $T$
halts and accepts an input string. Then, $Reify~accepts$ is a linear type that
captures precisely the string the same language as $T$. That is,
$Lang(Reify~accepts)$ is recursively enumerable --- the most general class of
languages in the Chomsky hierarchy.
\begin{construction}[\Agda]
  \label{cons:turing}
  For any Turing machine $T$, we can construct a grammar in \theoryabbv
  that accepts the same language as $T$.
\end{construction}

@@

Parse transformers can be composed: given two parse transformers $f$ and $g$,
their composition is defined pointwise, i.e. $(f\circ g)_w = f_w \circ g_w$.
Furthermore, given a formal grammar $A$, its identity transformer is $id_w =
id_{A(w)}$, where $id_{A(w)}$ is the identity function on the set $A(w)$. This
defines a \emph{category}.
\begin{definition}
  Define $\Grammar$ to be the category whose objects are formal
  grammars and morphisms are parse transformers.
\end{definition}
This category is equivalent to the slice category $\Set/\Sigma^*$ and
as such is very well-behaved. It is complete, co-complete, Cartesian
closed and carries a monoidal biclosed structure. We will use these
structures to model the linear types, terms and equalities in
\theoryabbv. These categorical properties are precisely what is
required to interpret all of the linear type and term constructors.

@@

The main difference
between our shallow embedding and \theoryabbv is that our linear terms
are written in a combinator-style, without being able to use named
variables in linear terms. For example, the function $h$ from
\cref{fig:kleeneabstractproof} would be written with the fold combinator applied
to subexpressions for the $nil$ and $cons$ cases,
\(
h = fold~nil~(cons \circ id \otimes cons \circ assoc^{-1})
\). Notice in the $cons$ case of the fold, we manually reassociate
$(\literal{a} \otimes \literal{a}) \otimes \literal{a}^{*}$ to
$\literal{a} \otimes (\literal{a} \otimes \literal{a}^{*})$, then act on the
left by the identity and on the right by $cons$ to get a parse
$\literal{a} \otimes \literal{a}^{*}$, and finally we act again by $cons$ to
finally produce a parse of $\literal{a}^{*}$.
A benefit of this shallow embedding is that
the parsers are immediately available to a larger Agda development, as
they are just normal Agda code. Although, even in the small example from
\cref{fig:kleeneabstractproof} we must manage additional complexity --- such as
manually reassociations --- and
this only grows more complex in larger programs. In future work we will look to
implement a type checker for a syntax closer to the presentation in
this paper, while maintaining the goal of integration with an
existing proof assistant.

>>>>

To show the feasibility of our design, we have implemented \theoryabbv
as a shallowly embedded domain-specific language in the Cubical Agda
proof assistant \cite{VezzosiMortbergAbel2019}. We have implemented many example
grammars and parsers in our system including regular expressions,
non-deterministic and deterministic automata, as well as some example
$\LL{1}$ context-free grammars and parsers using stack-based
automata. Throughout this paper \Agda~ will mark
results that are mechanized in our Agda development and provide a link to their
implementation.

@@

\begin{axiom}
  \label{ax:string-top}
   $\lambda s. \mathsf{read}(!(s)) \equiv \lambda
s. s$ where $!$ is the unique function $\ltonl(\StringGram\lto \top)$.
\end{axiom}

@@

\section{Formal Grammar Theory in \theoryname}
\label{sec:applications}
This section explores the applications of \theoryabbv to formal
grammar theory. We demonstrate that several classical notions and constructions
integral to the theory of formal languages are faithfully represented
in \theoryname.

@@

A formal grammar $\A$ is ambiguous if there are multiple parse trees
for the \emph{same} string $\w$.  For example, $\a \oplus \a$ is
ambiguous because there are two parses of $\stringquote{a}$,
constructed using $\inl$ and $\inr$. On the other hand, a formal
grammar is unambiguous when there is at most one parse tree for any
input string. We can capture this notion as a type in \theoryabbv as follows:
\begin{definition}
  \label{def:unambig}
  A grammar $\A$ is \emph{unambiguous} if for every linear type $\B$,
  $\f : \ltonl{(\B \lto \A)}$, and $\g : \ltonl{(\B \lto \A)}$
  then $\f \equiv \g$.
\end{definition}
\cref{def:unambig} can be read more intuitively as stating that $\A$
is unambiguous if there is at most one way to transform parses of any
other grammar $\B$ into parses of $\A$. This notion of an unambiguous
type is the analog for linear types of the definition of a (homotopy)
\emph{proposition} in the terminology of homotopy type
theory \cite{hottbook}. The most basic unambiguous types are $\top$ and
$0$, and in a system of classical logic all unambiguous types would
have to be equivalent to one of these, but with our axioms we can show
also that $\I$ and literals $\literal c$ are unambiguous. To see this, first, we
establish two useful properties of unambiguity.
\begin{lemma}[\Agda]
  \label{lem:retract}
  If $\B$ is unambiguous and $\A$ is a retract of $\B$ then $\A$ is unambiguous.
\end{lemma}
\begin{lemma}[\Agda]
  \label{lem:unambig-sum}
  As a consequence of \cref{cor:binary-mono}, if a binary disjunction $A \oplus B$ is unambiguous then
  $A$ and $B$ are each unambiguous.
\end{lemma}
From \cref{lem:retract}, we can prove that $\StringGram$ is unambiguous,
since it is a retract of $\top$. In fact, observe that if $\A$ is a
retract of $\B$ and $\B$ is unambiguous, then in fact $\A$ and $\B$
are strongly equivalent, as the equation $\lambda b. f(g(b)) \equiv
\lambda b. b$ follows because $\B$ is unambiguous. Therefore
$\StringGram$ is also strongly equivalent to $\top$. Next, since $\StringGram$ is
defined as a Kleene star, we can easily show that $\StringGram \cong \I
\oplus \CharGram \oplus (\CharGram \otimes \CharGram \otimes \StringGram)$. Then by
\cref{lem:unambig-sum}, we have that $\I$ and
$\CharGram$ are unambiguous as well. Using the finiteness of the alphabet
$\Sigma$ and the unambiguity of $\CharGram$,
we have that each literal $\literal c$ is likewise unambiguous.

@@

Next, we describe how to construct an intrinsically
verified parser for regular expressions by compiling it to an NFA and
then a DFA. That is, for each regular expression $A$, we construct an
NFA $N(A)$ and a corresponding DFA $D(A)$ such that $A$ is strongly
equivalent to the traces of $N(A)$ and weakly equivalent to the
accepting traces of $D(A)$. Then we can easily construct a parser for
traces of $D(A)$ and apply \cref{lem:wk-eqv-parse} to get
a verified regular expression parser.

@@

Because DFAs are deterministic, we are able to prove that their types
of traces are unambiguous and define a parser for them directly. In particular
we show that for any start state $s$, ${\LinSigTy {b} {\Bool}
  \Trace_{D}~s~b}$ is a retract of $\StringGram$ and apply \cref{lem:retract} to
derive unambiguity. That is, we first
construct a function $\pparse_{D}$ which is a parser for
$\Trace_D~{s}~{true}$ with $\Trace_D~{s}~{false}$ being the disjoint
type used, and disjointness follows from the
unambiguity of $\LinSigTy {b} {\Bool} {\Trace_{D}~s~b}$ by \cref{lem:unambig-sum}.

@@

We implement the parser for the Dyck language using an \emph{infinite} state deterministic
automaton, in \Cref{fig:DyckAutomaton}. Here the state is a ``stack''
counting how many open parentheses have been seen so far. Functions
$parse_{M}$ and $print_{M}$ for this automaton can be defined
analogously to those for DFAs, and so $\LinSigTy {s}
{M .states} {\LinSigTy {b}{Bool} {Trace_{M}~s~b}}$ is likewise
unambiguous.

@@

This reification operation on functions $\StringSem \to U$ is very expressive,
as it sidesteps our linear typing connectives and utilizes the whole of nonlinear
dependent type theory to define a grammar. For example, given a
Turing machine $T$ one may
define a non-linear predicate $accepts : \StringSem \to U$ that encodes that $T$
halts and accepts an input string. Then, $Reify~accepts$ is a linear type that
captures precisely the string the same language as $T$. In general,
$Lang(Reify~accepts)$ is recursively enumerable --- the most general class of
languages in the Chomsky hierarchy.
\begin{construction}[\Agda]
  \label{cons:turing}
  For any Turing machine $T$, we can construct a grammar in \theoryabbv
  that accepts the same language as $T$.
\end{construction}

@@

Parse transformers can be composed: given two parse transformers $f$ and $g$,
their composition is defined pointwise, i.e. $(f\circ g)_w = f_w \circ g_w$.
Furthermore, given a formal grammar $A$, its identity transformer is $id_w =
id_{A(w)}$, where $id_{A(w)}$ is the identity function on the set $A(w)$. This
defines a \emph{category}.
\begin{definition}
  Define $\Grammar$ to be the category whose objects are formal
  grammars and morphisms are parse transformers.
\end{definition}
This category is equivalent to the slice category $\Set/\Sigma^*$ and
is very well-behaved. It is complete, co-complete, Cartesian
closed and carries a monoidal biclosed structure. We will use these
structures to model the linear types, terms and equalities in
\theoryabbv.

@@

The main difference
between our shallow embedding and \theoryabbv is that our linear terms
are written in a combinator-style, without being able to use named
variables in linear terms. For example, the function $h$ from
\cref{fig:kleeneabstractproof} would be written with the fold combinator applied
to subexpressions for the $nil$ and $cons$ cases,
\(
h = fold~nil~(cons \circ id \otimes cons \circ assoc^{-1})
\). Notice in the $cons$ case of the fold, we manually reassociate
$(\literal{a} \otimes \literal{a}) \otimes \literal{a}^{*}$ to
$\literal{a} \otimes (\literal{a} \otimes \literal{a}^{*})$, then act on the
left by the identity and on the right by $cons$ to get a parse
$\literal{a} \otimes \literal{a}^{*}$, and finally we act again by $cons$ to
finally produce a parse of $\literal{a}^{*}$.
Even in the small example from
\cref{fig:kleeneabstractproof} we must manage additional complexity --- such as
manually reassociations --- and
this only grows more complex in larger programs. On the other hand, a benefit of this shallow embedding is that
the parsers are immediately available to a larger Agda development, as
they are just normal Agda code. In future work we will look to
implement a type checker for a syntax closer to the presentation in
this paper, while maintaining the easy integration with an
existing proof assistant.
